Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-30 parallel-tractability-J-v9/response.tex

\begin{quote}
Example 6 does need a bit of revision.  It is not the case that Oex3 cannot
be materialzed tractably in parallel.  Instead there has to be a class of
ontologies that cannot be so materialized.
\end{quote}

We did the revision for Example~6 (see the paragraph following Example~6),
and Example~1 (see the last paragraph of Section~3.2).

\begin{quote}
It would be much better if the ontology names corresponded to the example
numbers, i.e., the ontology in Example 6 was Oex6.
\end{quote}

We made the ontology subscripts correspond to the example numbers
for Example~5, Example~6, Example~7 and Example~9.

@@

...

\begin{quote}
The paper is all about parallel tractability of materialization, both
theoretical and practical.  However the paper gives insufficient attention
to RDFS, which is the main ontology language used in practice, and
particularly for large knowledge bases.  Even though RDFS is laughably
simple it does not have parallel tractability of even entailment.  The paper
should better examine parallel tractability in the context of RDFS or RDFS
with minor extensions.

Of course, RDFS does not fit into the approach of the paper, as it does not
a-priori separate classes and individuals.  The paper needs to discuss this
issue, as it prominently mentions RDFS.
\end{quote}

We used a new section (Section~4.4) in the revision to discuss why RDFS
does not have parallel tractability of entailment.

@@

\begin{quote}
The paper does a bad job of describing why materialization in RDFS is not
parallizable.  RDFS KBs have an infinite number of consequences, so it is
certainly not possible to materialize all of them quickly.  There are,
however, well known tricks that can be used to generate a finite
representation of this infinite set of consequences.  However, even this
cannot be done quickly in parallel.  In fact, even RDFS entailment cannot be
done quickly in parallel, because RDFS allows for a kind of twisted
interaction (which I think it similar to twisted paths in the paper) when
computing subproperty relationships.  Of course, having subproperties of
rdfs:subPropertyOf is not a usual thing to do in RDFS.  The paper needs to
be much more clear on its relationship to RDFS.
\end{quote}

We gave an example that RDFS entailment may lead to the situation of path twisting in Section~4.4 of the revision.
We also discussed in what cases RDFS has parallel tractability of entailment.

### 2018-01-30 parallel-tractability-J-v9/background.tex
Background proof-read

In what follows, we use $\textbf{CN}$, $\textbf{RN}$ and $\textbf{IN}$
to denote three disjoint countably
infinite sets of \emph{concept names, role names}, and \emph{individual names} respectively.
The set of roles is defined as $\textbf{R}:=\textbf{RN}\cup\{R^-|R\in\textbf{RN}\}$
where $R^-$ is the \emph{inverse role} of $R$.
For ease of discussion, we focus on the \emph{simple forms} of axioms shown
in the left column of Table~1. These simple forms can be obtained by using
well-known \emph{structure transformation} techniques \cite{KrotzschRH07,Kazakov09}.

DHL (short for \emph{description horn logic}) \cite{GrosofHVD03} is introduced as an
intersection of description logic (DL) and datalog in terms of expressivity.
We define a DHL ontology $\mathcal{O}$ as a triple:
$\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$, where
$\mathcal{T}$ denotes the TBox containing axioms of the forms (T1-T3);
$\mathcal{R}$ is the RBox that is a set of axioms of the forms (R1-R3);
$\mathcal{A}$ is the ABox containing \emph{assertions} of the forms (A1) and (A2).
In an axiom of either of  the forms (T1-T3 and R1-R3), concepts $A_{(i)}$ and $B$ are either
concept names, \emph{top concept} ($\top$) or \emph{bottom concept} ($\bot$); $R$ and $S_{(i)}$
are roles in $\textbf{R}$.
For an axiom of the form $A\sqsubseteq\forall R.B$ that is also allowed in DHL, we only consider its
equivalent form $\exists R^-.A\sqsubseteq B$.

DHL is related with other ontology languages.
First, DHL is essentially a fragment of the description logic Horn-$\mathcal{SHOIQ}$ with
disallowing \emph{nominal}, \emph{number restriction} and
right-hand \emph{existential restriction} ($A\sqsubseteq\exists R.B$).
Second, the expressivity of DHL covers that of RDFS to some extent \cite{GrosofHVD03}.
Reasoning with RDFS ontologies is \texttt{NP}-complete \cite{Horst05}
and, thus, is not parallelly tractable.
However, by applying some simplifications and restrictions, RDFS ontologies can be
expressed in DHL \cite{GrosofHVD03} that has \texttt{PTime}-complete complexity for
materialization.

@@

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed to datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of the form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

DL-Lite is a group of ontology languages designed for highly-efficient
query answering through knowledge
and underpins the OWL profile OWL QL \cite{CalvaneseGLLR07}.
DL-Lite$_{core}$ and DL-Lite$_{\mathcal{R}}$ are the two basic fragments
in DL-Lite.
DL-Lite$_{core}$ requires that an TBox contains axioms of the forms (T1) $A\sqsubseteq B$,
(T2) $A_1\sqcap A_2\sqsubseteq B$ where $B\equiv\bot$, (T3) $\exists R.A\sqsubseteq B$
where $A\equiv\top$ or $B\equiv\bot$, and (T4) $A\sqsubseteq\exists R$,
an ABox contains \emph{assertions} of the forms (A1) and (A2), and
RBox is not involved. DL-Lite$_{\mathcal{R}}$ is an extension of DL-Lite$_{core}$,
which allows involving RBoxes that contain axioms of the forms (R1-R2).
If not specially specified, \emph{a DL-Lite ontology denotes a DL-Lite$_{\mathcal{R}}$ ontology
in the following paragraphs}.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is called a \emph{free variable} (it is
not occurring in the head atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we also use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules transformed from the axioms in $\mathcal{T}$ and $\mathcal{R}$,
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which is initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

\subsection{The NC Complexity}

The parallel complexity class NC, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the NC class
can be solved in poly-logarithmic time on a PRAM (parallel, random-access machine) with
a polynomial number of processors. We also say that an NC problem can be solved
in \emph{parallel poly-logarithmic time}.
Although the NC complexity is a theoretical analysis tool,
it has been shown that many NC problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the NC problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The NC complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in NC.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

>>>>

In what follows, we use $\textbf{CN}$, $\textbf{RN}$ and $\textbf{IN}$
to denote three disjoint countably
infinite sets of \emph{concept names, role names}, and \emph{individual names} respectively.
The set of roles is defined as $\textbf{R}:=\textbf{RN}\cup\{R^-|R\in\textbf{RN}\}$
where $R^-$ is the \emph{inverse role} of $R$.
For ease of discussion, we focus on the \emph{simple forms} of axioms shown
in the left column of Table~1. These simple forms can be obtained by using
well-known \emph{structural transformation} techniques \cite{KrotzschRH07,Kazakov09}.

DHL (short for \emph{description horn logic}) \cite{GrosofHVD03} is introduced as an
intersection of description logic (DL) and datalog in terms of expressivity.
We define a DHL ontology $\mathcal{O}$ as a triple:
$\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$, where
$\mathcal{T}$ denotes the TBox containing axioms of the forms (T1-T3);
$\mathcal{R}$ is the RBox that is a set of axioms of the forms (R1-R3);
$\mathcal{A}$ is the ABox containing \emph{assertions} of the forms (A1) and (A2).
In an axiom of either of  the forms (T1-T3 and R1-R3), concepts $A_{(i)}$ and $B$ are either
concept names, the \emph{top concept} ($\top$) or the \emph{bottom concept} ($\bot$); $R$ and $S_{(i)}$
are roles in $\textbf{R}$.
For an axiom of the form $A\sqsubseteq\forall R.B$ that is also allowed in DHL, we only consider its
equivalent form $\exists R^-.A\sqsubseteq B$.

DHL is related with other ontology languages.
First, DHL is essentially a fragment of the description logic Horn-$\mathcal{SHOIQ}$ with
disallowing \emph{nominals}, \emph{number restrictions} and
right-hand side \emph{existential restrictions} ($A\sqsubseteq\exists R.B$).
Second, the expressivity of DHL covers that of RDFS to some extent \cite{GrosofHVD03}.
Reasoning with RDFS ontologies is \texttt{NP}-complete \cite{Horst05}
and, thus, is not tractable in parallel.
However, by applying some simplifications and restrictions, RDFS ontologies can be
expressed in DHL \cite{GrosofHVD03}, which has \texttt{PTime}-complete complexity for
materialization.

@@

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed into datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

DL-Lite is a group of ontology languages designed for highly-efficient
query answering over knowledge bases
and underpins the OWL profile OWL QL \cite{CalvaneseGLLR07}.
DL-Lite$_{\text{core}}$ and DL-Lite$_{\mathcal{R}}$ are the two basic fragments
of DL-Lite.
DL-Lite$_{\text{core}}$ requires that a TBox only contains axioms of the forms (T1) $A\sqsubseteq B$,
(T2) $A_1\sqcap A_2\sqsubseteq B$ where $B\equiv\bot$, (T3) $\exists R.A\sqsubseteq B$
where $A\equiv\top$ or $B\equiv\bot$, and (T4) $A\sqsubseteq\exists R$,
an ABox contains \emph{assertions} of the forms (A1) and (A2), and
RBoxes are not allowed. DL-Lite$_{\mathcal{R}}$ is an extension of DL-Lite$_{\text{core}}$,
which allows involving RBoxes that contain axioms of the forms (R1-R2).
If not specially specified, \emph{a DL-Lite ontology denotes a DL-Lite$_{\mathcal{R}}$ ontology
in the following paragraphs}.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{\text{core}}$ or DL-Lite$_{\mathcal{R}}$ can be transformed into a datalog program
(see the corresponding rules in the right column of Table~1).
Note that an axiom of form (T4) $A\sqsubseteq\exists R$ should be
transformed into a
first-order logic rule of the form $A(x)\rightarrow \exists y(R(x,y))$, where $y$ is called a \emph{free variable} (it is
not occurring in the body atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a fresh individual $o_{R}^A$
that corresponds to the concept $A$ and the role $R$. The elimination of free variables
does not influence the result of materialization \cite{CalvaneseGLLR07}.

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules obtained by transforming the axioms in
$\mathcal{T}$ and $\mathcal{R}$ and
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which was initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

\subsection{The Complexity Class \texttt{NC}}

The parallel complexity class \texttt{NC}, known as Nick's
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the \texttt{NC} class
can be solved in poly-logarithmic time on a PRAM (parallel, random-access machine) with
a polynomial number of processors. We also say that an \texttt{NC} problem can be solved
in \emph{parallel poly-logarithmic time}.
Although the \texttt{NC} complexity class is a theoretical analysis tool,
it has been shown that many \texttt{NC} problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, \texttt{NC} problems are also
highly feasible in parallel for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The \texttt{NC} complexity class was originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem is a decision problem in \texttt{NC}.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.
